Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-04-07 doc/manuscript_hkk_drought_sensitivity.Rmd
Update manuscript_hkk_drought_sensitivity.Rmd
refining intro & a bit in discussion. Just pushing to keep pushes small/easier to review

<!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
Tree species have diverse evolutionary strategies <!--that span and interact across timescales --->to cope with drought, often leading to their classification as "drought tolerant" or "drought sensitive" [@refs].
In tropical seasonal forests in particular, trees have evolved to maximise growth and survival under periodic hot/dry conditions, with strategies ranging along a spectrum from hydraulic safety to efficiency leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021]. 
Dry season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid water loss and have acquisitive strategies during leaf on periods [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf turgor loss point (hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), large vessels (hydraulic efficiency) [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021]. 
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls; e.g. some species regulate negative TLPs, leaf deciduousness and root depth, based on environmental cues leading to intraspecific or interannual variation potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; REFS Lasky? Hulshof?].

@@

Individual growth responses to drought can depend on the specific microenvironments that trees experience leading to varied atmospheric and hydrological stresses with distinct ecological consequences. 

@@

This results in greater stress on trees that have exposed crowns [@vinod_thermal_2023], shallower roots or growing in drier microenvironments []. 
Consequently, larger trees, with more exposed crowns, suffer more growth declines during drought [@bennett_larger_2015]<!--; larger trees have more exposed crowns although crown exposure can allow access to light and increase growth in normal years-->. 

@@

<!---CONCLUSIONS---->
We show strong evidence that tropical tree drought sensitivity can be jointly and interactively shaped by drought characteristics, species adaptations and local environmental conditions (Table 1).
Analysing two droughts in a 14-year dendrometer band timeseries, we found that tropical tree species and individuals grew less than average in drought years, reflecting in lower plot-level growth (Fig 2). 
Comparing across droughts, we found that despite the shorter duration of drought conditions, the wet-season drought elicited stronger responses across trees, but especially so for deciduous species that are expected to have stronger seasonality in growth phenology (Fig 3). 
Interestingly, in the wet-season drought, we also show that evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites (Fig 3b, Fig 4b). 
Further, besides consistent negative effects of crown exposure, we found that the drivers of variation among species and individuals in drought sensitivity were dissimilar between the drought events (Fig 4a). 
Water availability buffered exposure-associated growth declines through different mechanisms in the two different droughts - larger trees (with likely deeper roots) in the dry-season drought and trees in wetter sites in the wet-season drought suffered smaller declines in growth (Fig 5). 
Taken together, our analysis of tree growth in two different drought years, and spanning meaningful variation, shows that species and individuals can respond differently to different drought events. 
Despite stand-level growth declines in drought years, we show that these broad patterns belie enormous heterogeneity in individual and species responses to drought in tropical forests with possible consequences for the maintenance of species diveristy and ecosystem stability.

>>>>

<!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
Tree species have diverse evolutionary strategies <!--that span and interact across timescales --->to cope with dry conditions, often leading to their classification as "drought tolerant" or "drought sensitive" [@refs].
In tropical seasonal forests in particular, trees have evolved to maximise growth and survival under periodic hot/dry conditions, with strategies ranging along a spectrum from hydraulic safety to efficiency leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021]. 
Dry season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss and have acquisitive strategies during leaf-on periods [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and *large vessels (hydraulic efficiency)*<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021]. 
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls.
For example, some species regulate negative $\pi_{tlp}$s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation potentially and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; REFS Lasky? Hulshof?].

@@

Individual growth responses to drought can depend on the specific microenvironments that trees experience, leading to varied atmospheric and hydrological stresses with distinct ecological consequences. 

@@

This results in greater biophysical stress on trees that have exposed crowns [@vinod_thermal_2023] or rely on more ephemeral (typically shallower) water sources that are severely diminished under drought [@refs].<!--I (KAT) made a substantive change here. It's not necessarily the case that shallow-rooted trees or trees in dry microenvironments suffer more, as they'd also tend to have adaptations to handle those conditions.--> 
Consequently, larger trees, with more exposed crowns, suffer more growth declines during drought [@bennett_larger_2015] (**see issue [#24](https://github.com/forestgeo/growth-precip-thailand/issues/24)**)<!--; larger trees have more exposed crowns although crown exposure can allow access to light and increase growth in normal years-->. 

@@

<!---CONCLUSIONS---->
We show strong evidence that tropical tree drought sensitivity can be jointly and interactively shaped by drought characteristics, species adaptations and local environmental conditions (Table 1).
Analysing two droughts in a 14-year dendrometer band timeseries, we found that tropical tree species and individuals grew less than average in drought years, reflecting in lower plot-level growth (Fig 2). 
Comparing across droughts, we found that despite the *shorter duration of drought conditions*<!--careful here - didn't this continue into 2016?--> and greater absolute extremes of temperature, vapor pressure deficit, and lack of precipitation (Fig. 2a), the wet-season drought elicited stronger responses across trees, especially deciduous species that are expected to have stronger seasonality in growth phenology (Fig 3). 
Interestingly, in the wet-season drought, we also show that evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites (Fig 3b, Fig 4b). 
Further, besides consistent negative effects of crown exposure, we found that the drivers of variation among species and individuals in drought sensitivity were dissimilar between the drought events (Fig 4a). 
Water availability buffered exposure-associated growth declines through different mechanisms in the two different droughts - larger trees (with likely deeper roots) in the dry-season drought and trees in wetter sites in the wet-season drought suffered smaller declines in growth (Fig 5). 
Taken together, our analysis of tree growth in two different drought years, and spanning meaningful variation, shows that species and individuals can respond differently to different drought events. 
Despite stand-level growth declines in drought years, we show that these broad patterns belie enormous heterogeneity in individual and species responses to drought in tropical forests with possible consequences for the maintenance of species diveristy and ecosystem stability.
